$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-01 paper/paper.tex
structural congruence

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}
\label{syntax}
We let $x, y,\ldots$ range over \emph{channel names} or \emph{variables} and
$P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by the following grammar.
\begin{equation*}
  \begin{split}
    P,Q ::=      & \; \PO       &\text{(terminated process)}  \\
                |& \; (\new{}x)  P   & \text{(restriction)} \\
                |& \; \comp{P}{Q}  & \text{(parallel composition)} \\
                |& \; \recvp{x}{y}.P  & \text{(input)} \\
                |& \; \sendp{x}{y}.P   &\text{(output)}
  \end{split}
\end{equation*}
Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)P$ is the restriction process where channel $x$ is created new and is bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y}.P$ and $\sendp{x}{y}.P$ denote respectively the input and output processes of a variable/channel $y$ over a channel $x$, with continuation $P$.

@@

Using type-level de Bruijn indices makes our syntax well-scoped by construction --- the semantics of our language can now be defined on the totality of the syntax.
In order to make the de Bruijn indices representation more human-readable, we can annotate these indices with the original channel names.
Machines can then manipulate references mechanically whilst using channel names to present \picalc{} terms back to humans.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Semantics}
\label{semantics}

In the \lambdacalc{} $\beta$-reduction operates on syntactically adjacent terms.
In the \picalc{} however, the syntax introduces unnecessary distinctions
(e.g. semantically, parallel composition is defined modulo associativity and commutativity).
There are several ways around this, a structural congruence relation being one of the historical ones \cite{}.
(Others include labeled transition systems \cite{} and higher inductive types \cite{}.)

@@

Structural congruence is a congruent equivalence relation on processes.
Figure \ref{fig:struct-cong-base} lists the base cases of structural congruence.
The type $\Unused_i \; Q$ is an inductive proof that witnesses that the variable index $i$ does not appear nor in the inputs nor in the outputs of $Q$.
The function $lower_i \; Q \; uQ$ traverses $Q$ decrementing every index bigger than $i$.
Finally, $swap_i \; P$ traverses $P$ (of type $\Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all three, $i$ is incremented every time we go under a binder.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {P \eqeq Q : \Set}
  \; \textsc{StructCong}

@@

\begin{nilemma}
  \label{lm:lower-lift}
  The function $lower_i \; P \; uP$ has an inverse $lift_i \; P$ that increments every $\textsc{Var}$ greater than or equal to $i$, such that $lift_i \; (lower_i \; P \; uP) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nilemma}
  \label{lm:swap-swap}
  The function $swap_i \; P$ is its own inverse: $swap_i \; (swap_i \; P) \equiv P$.
\end{nilemma}
\begin{proof}
  By structural induction on \textsc{Process} and \textsc{Var}.
\end{proof}

Structural congruence is a congruent equivalence relation.
As such, rewrites can happen anywhere inside a process, and are closed under reflexivity, symmetry and transitivity as shown in Figure \ref{fig:struct-cong}.
In \S \ref{subject-congruence} we prove that if two processes $P$ and $Q$ are structurally congruent and $P$ is well-typed, then so is $Q$.
In the transitivity case, we must show that if $P$ is structurally congruent to $Q$ and $Q$ is to $R$, and $P$ is well-typed, then so is $R$.
To do so, we need to proceed by induction and first get a proof of the well-typedness of $Q$, then use that to reach $R$.
To show the typechecker that the doubly recursive call terminates we index the equivalence relation $=$ by a type $\Rec$ that models the structure of the recursion.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {\Rec : \Set}
  \; \textsc{Rec}

  \inferrule
  { }
  {\constr{zero} : \Rec}

  \inferrule
  {r : \Rec}
  {\constr{one} \; r : \Rec}

  \inferrule
  {r \; s : \Rec}
  {\constr{two} \; r \; s : \Rec}

  \datatype
  {P \, Q : \Process_n \\ r : \Rec}
  {P \eq{r} Q : \Set}
  \; \textsc{Equals}

  \inferrule
  {eq : P \eqeq Q}
  {\constr{struct} \; eq : P \eq{\constr{zero}} Q}

@@

  \inferrule
  {eq : P \eq{r} P'}
  {\constr{cong-send} \; eq : \send{x}{y}P \eq{\constr{one} \; r} \send{x}{y}P'}

  \inferrule
  { }
  {\constr{refl} : P \eq{\constr{zero}} P}

  \inferrule
  {eq_1 : P \eq{r} Q \\ \; eq_2 : Q \eq{s} R}
  {\constr{trans} \; eq_1 \; eq_2 : P \eq{\constr{two} \; r \; s} R}
\end{mathpar}
\caption{Structural rewriting rules lifted to a congruent equivalence relation indexed by a recursion tree.}
\label{fig:struct-cong}
\end{figure}

@@

\newpage
\bibliography{paper}
\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}
\label{syntax}
We let $x, y,\ldots$ range over \emph{channel names} or \emph{variables} and
$P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by the following grammar.
\todo{maybe put both syntaxes in two columns to save space}
\begin{equation*}
  \begin{split}
    P,Q ::=      & \; \PO       &\text{(terminated process)}  \\
                |& \; (\new{}x)  P   & \text{(restriction)} \\
                |& \; \comp{P}{Q}  & \text{(parallel composition)} \\
                |& \; \recvp{x}{y}.P  & \text{(input)} \\
                |& \; \sendp{x}{y}.P   &\text{(output)}
  \end{split}
\end{equation*}
Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)P$ is the restriction process where channel $x$ is created new and is bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y}.P$ and $\sendp{x}{y}.P$ denote respectively the input and output processes of a variable/channel $y$ over a channel $x$, with continuation $P$.

@@

Using type-level de Bruijn indices makes our syntax well-scoped by construction --- the semantics of our language can now be defined on the totality of the syntax.
In order to make the de Bruijn indices representation more human-readable, we can annotate these indices with the original channel names.
Machines can then manipulate references mechanically whilst using channel names to present \picalc{} processes back to humans.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Semantics}
\label{semantics}
We introduce the operational semantics for the \picalc{}, starting with structural congruence.

@@

The complete structural congruence is given in \autoref{app:struct}.
Figure \ref{fig:struct-cong-base} lists the base cases.

\todo{remove space between inferrules}
\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {P \eqeq Q : \Set}
  \; \textsc{StructCong}

@@

In $\constr{scope-ext}$ the type $\Unused_i \; Q$ is an inductive proof that witnesses that the variable index $i$ does not appear neither in the inputs nor in the outputs of $Q$.
The function $lower_i \; Q \; uQ$ traverses $Q$ decrementing every index bigger than $i$.
In $\constr{scope-comm}$ function $swap_i \; P$ traverses $P$ (of type $\Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all the above, $i$ is incremented every time we go under a binder.

Structural congruence is a congruent equivalence relation.
As such, rewrites can happen anywhere inside a process, and are closed under reflexivity, symmetry and transitivity as shown by the first row of Figure \ref{fig:struct-cong}.
The rest of the rules defines structural congruence under a context $\mathcal{C}[\cdot]$ \cite{Sangio01}, respectively restriction, composition, input and output.
%In \S \ref{subject-congruence} we prove that if two processes $P$ and $Q$ are structurally congruent and $P$ is well-typed, then so is $Q$.

\begin{figure}[h]
\begin{mathpar}
  \inferrule
  { }
  {\constr{refl} : P \eq{\constr{zero}} P}

  \inferrule
  {eq_1 : P \eq{r} Q \\ \; eq_2 : Q \eq{s} R}
  {\constr{trans} \; eq_1 \; eq_2 : P \eq{\constr{two} \; r \; s} R}

@@

  \inferrule
  {eq : P \eq{r} P'}
  {\constr{cong-send} \; eq : \send{x}{y}P \eq{\constr{one} \; r} \send{x}{y}P'}
\end{mathpar}
\caption{Structural congruence under contexts.}
\label{fig:struct-cong}
\end{figure}

@@

\newpage
\bibliography{paper}

%%
\newpage
\appendix
\input{appendix.tex}
\end{document}
